---
title: "COVID-19 United States Excess Deaths by county and month"
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include=FALSE}
library(tidyverse)
library(ggh4x)
library(data.table)
library(lubridate)
library(equatiomatic)
library(knitr)
library(tinytex)
library(texPreview)

knitr::opts_chunk$set(
  echo = FALSE,
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  dev.args = list(png = list(type = "cairo")), dpi = 96,
  fig.path = "README_files/"
)

options(
  scipen = 999,
  digits = 3,
  knitr.kable.NA = "n/a"
)
```


# Overview

```{r import}
lmer_model <- readRDS(
  file = file.path(
    here::here(),
    "results/united_states_county_monthly_model.RDS"
  )
)

united_states_county_monthly_excess_deaths_estimates <- data.table::fread(
  file.path(
    here::here(),
    "results/united_states_county_monthly_excess_deaths_estimates.csv"
  ),
  keepLeadingZeros = TRUE
)

total_excess_deaths_2020 <- united_states_county_monthly_excess_deaths_estimates %>%
  filter(year == 2020L, start_date >= "2020-03-01") %>%
  pull(excess_deaths) %>%
  sum(na.rm = TRUE)

total_excess_deaths_2021 <- united_states_county_monthly_excess_deaths_estimates %>%
  filter(year == 2021L, start_date < "2021-12-01") %>%
  pull(excess_deaths) %>%
  sum(na.rm = TRUE)
```

This repository contains code and data to estimate **expected deaths** and **excess deaths** in the United States in 2020 and 2021 by **county** and **month**.
This model estimates that there were about `r round(total_excess_deaths_2020)` excess deaths in the United States in 2020 (excluding January and February) and `r round(total_excess_deaths_2021)` in 2021 (excluding December), for a total of `r round(sum(total_excess_deaths_2020, total_excess_deaths_2021))`.

Estimated excess deaths are available at [`/blob/main/results/united_states_county_monthly_excess_deaths_estimates.csv`](https://github.com/mymil/covid-19-united-states-county-monthly-excess-deaths/blob/main/results/united_states_county_monthly_excess_deaths_estimates.csv).


#### Nationwide United States Mortality, 2011 through 2021 (provisional and partial)
```{r results summary, message=FALSE, warning=FALSE}

deaths_nationwide_summary <- united_states_county_monthly_excess_deaths_estimates %>% 
  group_by(start_date) %>% 
  summarise(
    across(c(total_deaths, expected_deaths, excess_deaths), sum, na.rm = TRUE),
    .groups = "drop"
  ) %>% 
  mutate(period = date(start_date)) %>% 
  # exclude the prior month from plot due to partial data
  filter(
    period < floor_date(
      Sys.Date() - months(1), 
      unit = "month"
    )
  )

g <- ggplot(deaths_nationwide_summary, aes(x = period))

p <- g +
  geom_line(
    aes(y = expected_deaths, color = "Expected deaths")
  ) +
  geom_line(
    aes(y = total_deaths, color = "Total deaths")
  ) +
  stat_difference(
    aes(ymin = expected_deaths, ymax = total_deaths),
    levels = c("Excess mortality", NA, NA),
    alpha = .7
  ) +
  geom_vline(
    xintercept = date("2020-03-01"),
    linetype = "dashed"
  ) +
  annotate(
    geom = "curve",
    x = date("2018-10-01"),
    y = 310000,
    xend = date("2020-02-20"),
    yend = 310000,
    curvature = -.2,
    arrow = arrow(length = unit(3, "mm"))
  ) +
  annotate(
    geom = "text",
    x = date("2018-09-20"),
    y = 310000,
    label = "March 2020, the beginning of the\nCOVID-19 pandemic in the United States",
    hjust = "right"
  ) +
  # annotate(
  #   geom = "curve", 
  #   x = date("2019-12-01"), 
  #   y = 630000, 
  #   xend = date("2021-04-01"), 
  #   yend = 720000, 
  #   curvature = .4, 
  #   arrow = arrow(length = unit(3, "mm"))
  # ) +
  # annotate(
  #   geom = "text", 
  #   x = date("2019-11-20"), 
  #   y = 620000, 
  #   label = "Recent provisional mortality data\nare underreported due to reporting lags",
  #   hjust = "right"
  # ) +
  scale_color_viridis_d(option = "cividis", end = .6, direction = -1) +
  scale_fill_viridis_d(option = "cividis", na.translate = FALSE) +
  scale_x_date(
    date_breaks = "1 year",
    date_minor_breaks = "3 months",
    labels = scales::label_date_short()
  ) +
  guides(
    fill = guide_legend(NULL),
    color = guide_legend(NULL)
  ) +
  labs(
    x = "Month",
    y = "Deaths"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank()
  )

print(p)
```

Excess mortality for the months in which total deaths exceeded expected deaths are highlighted in blue.

# Data Sources

* CDC WONDER all-cause, all-ages deaths by county and month

  https://wonder.cdc.gov/ucd-icd10.html
  
  These public-use historical mortality data were obtained for years 2011-2020 and used to estimate the expected deaths model.
  2020 data were also compared against estimated expected deaths to calculate excess deaths.
  Because these data have an unusual and specific license, they are not included as part of the repository, but can be freely downloaded from https://wonder.cdc.gov/ucd-icd10.html.
  
  Note: County-months wither fewer than 10 deaths are censored in the source data.

* CDC WONDER provisional and partial all-cause deaths and deaths involving COVID-19 by county and month

  https://wonder.cdc.gov/mcd-icd10-provisional.html
  
  Example query: https://wonder.cdc.gov/controller/saved/D176/D273F449
  
  These public-use provisional and partial mortality data were obtained for 2021 and and compared to estimated expected deaths to calculate excess deaths.
  Because these data have an unusual and specific license, they are not included as part of the repository, but can be freely downloaded from https://wonder.cdc.gov/ucd-icd10.html.
  
  Note: Values for county-months with fewer than 10 deaths are suppressed in these data.
  
* US Census Population Estimates

  Population estimates for 2011-2020 were retrieved from the [US Census Bureau Population Estimates](https://www.census.gov/programs-surveys/popest/technical-documentation/research/evaluation-estimates/2020-evaluation-estimates/2010s-counties-total.html), because these data are not yet included in the Census Bureau Data API.
  2020 population estimates were used for 2021.
  
  Note: This product uses Census Bureau data but is not endorsed or certified by the Census Bureau.
  
# Excess Deaths Model

  The excess deaths model used began as an adaptation of *The Economist*'s [excess mortality model](https://github.com/TheEconomist/covid-19-excess-deaths-tracker), but has since diverged.
  For details on model selection and performance, see [`/blob/main/results/README.md`](https://github.com/mymil/covid-19-united-states-county-monthly-excess-deaths/blob/main/results/README.md).
  
  Given the large number of counties in the United States (over 3000), a linear mixed model with county, county set, and census division as random grouping factors was used to make estimation tractable.
  These random grouping factors enable each county, county set, and state to have their own intercepts in the final model.
  Furthermore, because mortality rates are correlated over time, an auto-regressive term was included in the model to allow mortality rates to correlate month-to-month within each county.
  
  More precisely, **total deaths per day** was regressed on:
  
  * county population (z-scored)
  
  * years since 2011
  
  * month of the year (fixed grouping factor)
  
  * county (random grouping factor nested within county set)
  
  * county set (random grouping factor nested within state)
  
  * state (random grouping factor)
  
  This model can be expressed by the equation: (NOTE: The equation is temporarily unavailable due to {equatiomatic} not currently having a method to convert glmmTMB models to LaTeX.)

```{r equation, message=FALSE}
# equatiomatic::extract_eq(
#   lmer_model,
#   wrap = TRUE,
#   terms_per_line = 2,
#   swap_var_names = c(
#     "total_deaths_per_day" = "Total Deaths per Day",
#     "population_z" = "County Population (z-score)",
#     "year_zero" = "Years since 2015",
#     "month" = "Month",
#     "region_code" = "County",
#     "county_set_code" = "County Set",
#     "state" = "State"
#   )
# ) %>%
#   texPreview::tex_preview(
#     stem = "plot_equation-1",
#     fileDir = knitr::opts_chunk$get()$fig.path,
#     returnType = "html"
#   )
```

The model object is available at [`/blob/main/results/united_states_county_monthly_model.RDS`](https://github.com/mymil/covid-19-united-states-county-monthly-excess-deaths/blob/main/results/united_states_county_monthly_model.RDS).

The model estimates for each observation, including fitted values and residuals, are available at [`/blob/main/results/united_states_county_monthly_fitted_deaths_per_day_estimate.csv`](https://github.com/mymil/covid-19-united-states-county-monthly-excess-deaths/blob/main/results/united_states_county_monthly_fitted_deaths_per_day_estimate.csv).
Note that these estimates are only y-hat directly from the initial model, **not** predicted values on the full dataset, which are [available here](https://github.com/mymil/covid-19-united-states-county-monthly-excess-deaths/blob/main/results/united_states_county_monthly_excess_deaths_estimates.csv)..

# Model Performance

For more details on model selection and performance, see [`/blob/main/results/README.md`](https://github.com/mymil/covid-19-united-states-county-monthly-excess-deaths/blob/main/results/README.md).

Because the COVID-19 pandemic only began in the United States partway through March, 2020, we can evaluate model performance by examining concordance of predicted and observed deaths in the beginning of 2020.

```{r performance}
predict_check <- united_states_county_monthly_excess_deaths_estimates %>%
  filter(start_date >= "2020-01-01" & start_date < "2020-04-01") %>% 
  rename(period = start_date)

result_corr <- cor.test(x = predict_check$total_deaths, y = predict_check$expected_deaths)
```

Observed and expected deaths in January through March, 2020, are highly correlated, r = `r result_corr$estimate`.
As can be seen from the following scatterplot, total deaths (y-axis) tended to exceed expected deaths (x-axis).

```{r plot_comparison, message=FALSE, warning=FALSE}
g <- ggplot(
  predict_check,
  aes(
    x = expected_deaths,
    y = total_deaths
  )
)

p <- g +
  geom_smooth(
    method = "lm",
    se = FALSE,
    fullrange = TRUE,
    color = "red",
    size = .3,
    linetype = "dotted"
  ) +
  geom_point(
    shape = 21,
    color = "grey20",
    alpha = .6
  ) +
  geom_abline(
    intercept = 0,
    slope = 1,
    size = 0.3,
    color = "grey25"
  ) +
  facet_wrap(vars(period)) +
  ggrepel::geom_label_repel(
    data = filter(
      predict_check,
      total_deaths > 2000 |
        (total_deaths / expected_deaths > 1.2 &
           total_deaths > 1000)
    ),
    aes(label = region),
    color = "grey10",
    segment.color = "grey40",
    size = 3,
    min.segment.length = 0,
    force = 10,
    force_pull = .0001,
    point.padding = .4,
    label.padding = .1,
    max.time = 5,
    label.size = NA,
    nudge_x = -2000,
    nudge_y = 1000,
    seed = 20211004
  ) +
  colorspace::scale_fill_continuous_diverging(palette = "cork") +
  coord_equal() +
  theme_minimal()

print(p)
```

# Repository Organization

* Excess deaths estimates and fitted LMM are available in `/results/`.

* Code to fit the model and estimate excess deaths are available in `/code/`.

* Data used in modeling are available in `/data/`, when possible.

# License

The code contained in this repository are available under the [MIT License](https://opensource.org/licenses/MIT), and the data generated by this code are licensed under the [Creative Commons Attribution 4.0 International License](https://creativecommons.org/licenses/by/4.0/).

```{r, echo=TRUE}
sessionInfo()
```
